---
title: "R Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data

We use Connecticut as an example because it is a small dataset and easy to load in but still has fairly complete data. 

```{r, message=FALSE,  warning=FALSE}
library(tidyverse)
library(lubridate)

d = read_csv('CT-clean.csv', col_types = list(stop_time = 'c', officer_id = 'c'))
```

## Data exploration

Now let's apply the same filters we used in the analysis.  We analyze stops of white, black, and Hispanic drivers between 2011 - 2015. 

```{r}
d = filter(d, 
           driver_race %in% c('White', 'Black', 'Hispanic'), 
           year(stop_date) >= 2011, 
           year(stop_date) <= 2015)
```

After filtering we lose about 8,000 rows. Now let's compute some basic statistics broken down by the race of the driver.

```{r}
summary_stats <- function(search_conducted, contraband_found) {
  n_stops     = length(search_conducted)
  n_searches  = sum(search_conducted)
  n_hits      = sum(contraband_found)
  search_rate = n_searches / n_stops
  hit_rate    = n_hits / n_searches
  return(data.frame(n_stops, n_searches, n_hits, search_rate, hit_rate))
}

basic_summary_statistics_by_race = d %>% 
  group_by(driver_race) %>% 
  do(summary_stats(.$search_conducted, .$contraband_found))
basic_summary_statistics_by_race
```

The first thing we notice is that black and Hispanic drivers are much more likely to be searched: 
the search rate is about 2 - 3x higher than for white drivers. This is a pattern that appears across states. 
This disparity, on its own, does not prove that the police are being discriminatory -- perhaps white drivers are 
less likely to carry contraband -- but is still worth noting.

Another thing we note is that the hit rate is higher for white drivers than for black and Hispanic drivers. 
This is the classic "outcome test": if searches of minority drivers are less likely to be successful, that may indicate
that minority drivers are searched when less likely to be carrying contraband, suggesting discriminatory search standards.
In general, the combination of higher search rates for minority drivers combined with lower hit rates suggests that
minority drivers are being searched on less evidence. 

Incidentally, race isn't the only thing you could stratify by. We can easily break down the data by gender or by other categories too. 
far fewer female drivers are stopped, and they're less likely to be searched, but have comparable hit rates. 

```{r}
basic_summary_statistics_by_gender = d %>% 
  group_by(driver_gender) %>% 
  do(summary_stats(.$search_conducted, .$contraband_found))
basic_summary_statistics_by_gender
```

We can also break things down by both race and location.
This is important to do because search rates or hit rates might vary by location for legitimate reasons
and racial composition could as well. So we want to see whether search and hit rates differ by race even when
we control for location. 

```{r}
basic_summary_statistics_by_race_and_county = d %>% 
  filter(!is.na(county_name)) %>%
  group_by(driver_race, county_name) %>%
  do(summary_stats(.$search_conducted, .$contraband_found))
```

Let's make a scatterplot which compares search rates and hit rates for minority and white drivers within the same county. 
```{r, warning=FALSE}
data_for_plot <- basic_summary_statistics_by_race_and_county %>%
  filter(driver_race == 'White') %>% 
  right_join(basic_summary_statistics_by_race_and_county %>% filter(driver_race != 'White'), by='county_name')

# plot search rates. 
max_val = max(basic_summary_statistics_by_race_and_county$search_rate) * 1.05
search_plot = ggplot(data_for_plot) + 
  geom_point(aes(x = search_rate.x, y = search_rate.y, size = n_stops.y)) + # specify data we want to plot
  facet_grid(.~driver_race.y) +   # make one subplot for each minority race group
  geom_abline(slope = 1, intercept = 0, linetype='dashed') +   # add a diagonal line to indicate parity
  scale_x_continuous('White search rate', limits=c(0, max_val), labels = scales::percent, expand=c(0,0)) + 
  scale_y_continuous('Minority search rate', limits=c(0, max_val), labels = scales::percent, expand=c(0,0)) + 
  theme_bw(base_size=15) + 
  theme(legend.position="none") + 
  scale_size_area(max_size=5)
search_plot
```

Points are all above the diagonal line, indicating that search rates are higher for minorities within same county.

```{r,  warning=FALSE}
max_val = max(basic_summary_statistics_by_race_and_county$hit_rate) * 1.05
hit_plot = ggplot(data_for_plot) + 
  geom_point(aes(x = hit_rate.x, y = hit_rate.y, size = n_stops.y)) + 
  facet_grid(.~driver_race.y) + 
  geom_abline(slope = 1, intercept = 0) + 
  scale_x_continuous('White hit rate', limits=c(0, max_val), labels = scales::percent, expand=c(0,0)) + 
  scale_y_continuous('Minority hit rate', limits=c(0, max_val), labels = scales::percent, expand=c(0,0)) + 
  theme_bw(base_size=15) + 
  theme(legend.position="none") + 
  scale_size_area(max_size=5)
hit_plot
```

Points are generally below the diagonal line, indicating that hit rates are lower for minorities within same county.

The outcome test isn't perfect because of what's known as the problem of "infra-marginality". 
We could observe different hit rates for different race groups even if there's no discrimination. 
Imagine there are two types of white drivers: those who have a 1% chance of carrying 
contraband, and those who have a 75% chance. Assume that black drivers have either a 1% or 50% chance of carrying contraband. 
Even if the police apply the same threshold, hit rates will be different!
You can read more about this problem, and solutions to it, in "The Problem of Infra-marginality in Outcome Tests for Discrimination". 

Another thing you can do is perform regressions. So maybe we want to look at search rates controlling for race, location, and age. 
```{r}
summary_stats_for_regression = d %>% 
  mutate(driver_age_category = cut(driver_age, 
                                   c(15, 19, 29, 39, 49, 100), 
                                   labels = c('16-19', '20-29', '30-39', '40-49', '50+')),
         driver_race = factor(driver_race, levels = c('White', 'Black', 'Hispanic'))) %>%
  group_by(driver_race, driver_age_category, driver_gender, county_name) %>% 
  do(summary_stats(.$search_conducted, .$contraband_found))
model = glm(cbind(n_searches, n_stops - n_searches) ~ driver_race + driver_age_category + driver_gender + county_name, data = summary_stats_for_regression, family = binomial)
summary(model)
```

So that's how to perform basic analyses for a single state. Of course, there's a lot of other things you could do!
There are many columns we haven't even looked at.

What if you want to scale up and analyze multiple states? You'll find that a lot of states take a lot longer to load in than 
Connecticut, and loading in all states will require a long time. We suggest working with aggregate data if you want to analyze all states.

